---
title: "clustering"
author: "Fang Liu"
date: "6/10/2022"
output: html_document
---

> This 'clustering.Rmd' 

First, load the required libraries 
```{r setup, message=FALSE}
library(readxl)
library(kohonen)
library(tidyverse)
library(factoextra)
library(RColorBrewer)

library(tidyverse)
library(caret)
library(rpart.plot) # for creating classification tree outputs
library(glmnet)
library(Amelia)
library(pROC) #for calculating ROC/AUC
library(gbm)
library(stats)
library(factoextra)
library(cluster)
```

Next, load in the data
```{r load_data, warning=FALSE, message=FALSE}

tox21 <- read.delim("./data/tox21_data.txt")
```

## Data Cleaning
Change rows with 'x' to NA and filter out drugs/chemicals with incomplete data using the **complete.case()** which returns a logical vector indicating which cases don't have missing values & returns ALL columns of the original data. Note that the # of rows dropped from 8971 to 7170.
```{r data_wrangling}
#replicate data and change rows with 'x' to NA 
new_df <- tox21
new_df[new_df == 'x'] <- NA  
data_filtered <- new_df[complete.cases(new_df),] 

#all columns are character, need to convert all columns to numerical
data_numerical <- as.data.frame(sapply(data_filtered[-c(1,2)], as.numeric)) 

#rejoin the CAS ID and bioassay data into a data frame 
final_data =
  data.frame(data_filtered$Structure.ID, data_numerical) %>% 
  janitor::clean_names() %>% 
  rename(structure_id = data_filtered_structure_id) 

#write to output
#write.csv(final_data, file = "results/tox21_data_cleaned.csv", row.names = FALSE)
#test <- read_csv("./results/tox21_data_cleaned.csv")
```

## SOM 
### Scaling 
SOM requires that ALL numerical values be scaled; thus, scale the data and make sure that the mean of each variable is 0.
```{r scale_data, results='hide'}
scaled_data <- scale(final_data[, -1]) #scale all columns except the first one 
summary(scaled_data) #check if scaling was successful(mean = zero)
```

### Create Grid 
Next, we can create and specify the grid to use for SOM.
```{r create_grid, results='hide'}
set.seed(123) #ensures reproducibility 

#create grid 
g <- somgrid(xdim = 12, ydim = 12, topo = "hexagonal", neighbourhood.fct = "gaussian")

#create map
map <- som(scaled_data, grid = g, alpha = c(0.05, 0.01), radius = 1, rlen = 100)
```

Plot our maps to see results.
```{r plot}
#reverse color (red = more)
colors <- function(n, alpha = 1) {
  rev(heat.colors(n, alpha))
}

plot(map, type = "codes") #shows the code book vectors; default
plot(map, type = "counts", palette.name = colors, heatkey = TRUE) #shows the # of objects mapped to the individual units

plot(map, type = "mapping", pchs = 20)
#symbols to plot when type equals "mapping", pchs = 20 means a filled circle 
plot(map, type = "dist.neighbours", palette.name = colors) #mapping distance between nodes; U-matrix?

#add cluster boundaries
#data.hc <- cutree(hclust(object.distances(map, "codes")), 2)
#add.cluster.boundaries(map, data.hc)
```


## K-means Clustering
```{r, eval=FALSE}
set.seed(100)

# Center and scale data
set.up.preprocess<-preProcess(final_data, method=c("center", "scale"))
transformed.vals<-predict(set.up.preprocess, final_data) %>% select(-1) #drop the CASRNs

# Conduct a gap statistic analysis to determine optimal number of clusters 
gap_stat<-clusGap(transformed.vals, FUN=kmeans, nstart=10, K.max=2, B=10, iter.max=50) 
print(gap_stat, method="firstmax")
fviz_gap_stat(gap_stat) 
```

*** 

#Create heatmaps
```{r}
scaled_data_w_cluster <- cbind(scaled_data, as.data.frame(map$unit.classif)) 
cluster_assignment_scaled <- scaled_data_w_cluster %>%
  rename(cluster = 'map$unit.classif') %>% 
  arrange(cluster)

new_data = cluster_assignment_scaled %>% 
  filter(cluster %in% c(9)) %>% 
  select(-cluster) 

data_matrix = as.matrix(new_data)

color <- colorRampPalette(brewer.pal(9, name = "RdBu"))(243) #BuPu

heatmap(data_matrix, cexRow=0.8, cexCol = 0.25, col = color, Colv = NA, Rowv = NA)

#Colv = NA, Rowv = NA

#heatmap(data_matrix,Rowv = NA, Colv = NA, cexRow=1, cexCol = 0.40)   

#write.csv(colnames(new_data), file = "results/clusters_names.csv")
```
