---
title: "enriched pathway analysis"
author: "Chloe Chen"
date: "12/16/2022"
output: html_document
---

> This `enriched_pathway.Rmd` file uses compound-gene relationships (pulled from BioPlanet) to identify a list of enriched pathway targets for each cluster. The code was modified from gene_enrichment_analysis.Rmd by Fang Liu. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(ClassComparison) #for 'Bum'
```

# Load pathway data
```{r}
pathway_data = read.csv("./data/pathway.csv")
```
# Add cluster information onto pathway data
```{r}
# read gene enrichment analysis results
gene_data = read.csv("./results/enrichment_analysis_results_7_20.csv") %>%
  filter(significant == TRUE) %>% #filter out insignificant genes
  select(cluster, target_gene) %>%
  group_by(cluster)
colnames(gene_data)[2] = "GENE_SYMBOL"

# add cluster information on pathway data
pathway_data_cluster = merge(pathway_data, gene_data, by="GENE_SYMBOL", all.x=TRUE)
pathway_data_cluster = na.omit(pathway_data_cluster)


length(unique(pathway_data_cluster$PATHWAY_NAME)) #1299 unique pathways
length(unique(pathway_data_cluster$GENE_SYMBOL)) #682 unique genes

```

# Overall pathway frequency 
```{r}
pathway_freq = pathway_data_cluster %>% 
  group_by(PATHWAY_NAME) %>% 
  summarize(n_obs = n()) %>% 
  arrange(-n_obs) 
```

# number of genes in each cluster 
```{r}
gene_cluster_freq = gene_data %>%
  group_by(cluster) %>% 
  summarize(cluster_size = n())

```

# find pathway frequency for each cluster
```{r}

pathway_within_cluster = 
pathway_data_cluster %>% 
  group_by(cluster, PATHWAY_NAME) %>% 
  summarize(n_pathways = n())

length(unique(pathway_within_cluster$PATHWAY_NAME)) #1299 unique pathway targets
```

# create final dataset 
```{r}
cluster_symbol_combined = 
  left_join(pathway_within_cluster, gene_cluster_freq, by = "cluster")

#add overall gene frequency data 
data_final =
  left_join(cluster_symbol_combined, pathway_freq, by = "PATHWAY_NAME") %>% 
  rename(num_gene_with_pathway_in_cluster = n_pathways,
         num_gene_in_cluster = cluster_size,
         num_gene_with_pathway = n_obs) 

#length(unique(data_final$target_gene))

data_fin = data_final %>% 
  mutate(TruePos = num_gene_with_pathway_in_cluster,
         FalsePos = num_gene_with_pathway - TruePos,
         FalseNeg = num_gene_in_cluster - TruePos,
         TrueNeg = 1299 - (TruePos + FalsePos + FalseNeg),
         TotalCmpdsTested = TruePos + FalsePos + FalseNeg + TrueNeg) %>% 
  select(cluster, PATHWAY_NAME, TruePos:TotalCmpdsTested) %>% 
  mutate(OddsRatio = (TruePos * TrueNeg)/(FalseNeg * FalsePos))

data_fin = na.omit(data_fin)

#find p-values 
p_val = apply(data_fin[, 3:6], 1, function(x) fisher.test(matrix(x, ncol=2),alternative = "greater")$p.value)

#add p-value column
data_fin$p_val <- p_val
```


# account for multiple comparison using the "BUM" class
```{r}
#create "bum" object using the vector of p-values (generated from fisher's exact test)
a <- Bum(p_val)

#graph distribution of p-values
hist(a)


# 0.0003966139
#use False Discovery Rate (FDR) to find the cut-ff value (set alpha = 0.01)
cutoffSignificant(a, alpha=0.0005,by='FDR') 

significant <- selectSignificant(a, alpha=0.0005, by='FDR') 
countSignificant(a, 0.0005, by='FDR') #number of significant p-values

#add significant column to results
data_fin$significant <- significant
```

```{r}
#output the results
write.csv(data_fin, "./results/pathway_enrichment_results.csv")
```

```{r}
all_enriched_pathways = data_fin %>%
  filter(significant == TRUE) %>%
  select(cluster, PATHWAY_NAME, p_val) %>%
  group_by(cluster) %>%
  arrange(cluster, p_val)
```
