---
title: "enrichment_analysis"
author: "Fang Liu"
date: "5/10/2022"
output: html_document
---

> This `enrichment_analysis.Rmd` file uses data from the drug repurposing hub and pharos to identify enriched gene targets for each cluster. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(ClassComparison) #for 'Bum'
```

# Load the combined pharos & drug hub data
```{r}
pharos_and_hub = read.csv("./results/inchikey/pharos_and_drughub.csv") %>% 
  select(cas, cluster, target_gene)

length(unique(pharos_and_hub$target_gene)) #1629 unique genes
length(unique(pharos_and_hub$cas)) #1829 unique compounds
```

# Overall gene frequency 
```{r}
gene_freq = pharos_and_hub %>% 
  group_by(target_gene) %>% 
  summarize(n_obs = n()) %>% 
  arrange(-n_obs) 
```

# number of compounds (with gene target) in each cluster 
```{r}
#compound_cluster_freq = read.csv("./clusters_results.csv") %>% 
#  group_by(cluster) %>% 
#  summarize(cluster_size = n())

compound_cluster_freq = pharos_and_hub %>% 
  select(-target_gene) %>% 
  unique() %>% 
  group_by(cluster) %>% 
  summarize(cluster_size = n())

#test if the above is right --> YES, WORKS!
#pharos_and_hub %>% filter(cluster == 8) %>% select(-target_gene) %>% unique() %>% count()

#sum(compound_cluster_freq$cluster_size) #1101
```

# find gene frequency for each cluster
```{r}
#columns: cas, cluster, target_gene

gene_within_cluster = 
pharos_and_hub %>% 
  group_by(cluster, target_gene) %>% 
  summarize(n_genes = n())

#length(unique(gene_within_cluster$target_gene))
```

# create final dataset 
```{r}
cluster_symbol_combined = 
  left_join(gene_within_cluster, compound_cluster_freq, by = "cluster")

#add overall gene frequency data 
data_final =
  left_join(cluster_symbol_combined, gene_freq, by = "target_gene") %>% 
  rename(num_compound_with_gene_in_cluster = n_genes,
         num_compound_in_cluster = cluster_size,
         num_compound_with_gene = n_obs) 

#length(unique(data_final$target_gene))#1303
#1101 --> 1829

data_fin = data_final %>% 
  mutate(TruePos = num_compound_with_gene_in_cluster,
         FalsePos = num_compound_with_gene - TruePos,
         FalseNeg = num_compound_in_cluster - TruePos,
         TrueNeg = 1829 - (TruePos + FalsePos + FalseNeg),
         TotalCmpdsTested = TruePos + FalsePos + FalseNeg + TrueNeg) %>% 
  select(cluster, target_gene, TruePos:TotalCmpdsTested) %>% 
  mutate(OddsRatio = (TruePos * TrueNeg)/(FalseNeg * FalsePos))

#find p-values 
p_val = apply(data_fin[, 3:6], 1, function(x) fisher.test(matrix(x, ncol=2),alternative = "greater")$p.value)

#head(p_val,1)

#Alternative hypothesis: true odds ratio is greater than 1 (i.e., compounds within the cluster has higher odds of having the gene than compounds outside of the cluster)
#if p-val < 0.05, than the above might be true! 

#add p-value column
data_fin$p_val <- p_val
```


# account for multiple comparison using the "BUM" class -> choose p-cutoff value
```{r}
#create "bum" object using the vector of p-values (generated from fisher's exact test)
a <- Bum(p_val)

#graph distribution of p-values
hist(a)

#use False Discovery Rate (FDR) to find the cut-ff value (set alpha = 0.01)
cutoffSignificant(a, alpha=0.01,by='FDR') 

#If p-value is less than 0.008570463, than it is significant.
significant <- selectSignificant(a, alpha=0.01, by='FDR') 
countSignificant(a, 0.05, by='FDR')
countSignificant(a, 0.01, by='FDR') #1050

#add significant column to results
data_fin$significant <- significant

# Create output file
#write.csv(data_fin, "~/Downloads/enrichment_analysis_results_4_28.csv")
```


# Add gene names
[HGNC](https://www.genenames.org/download/custom/)

```{r}
#read gene names... 
hgnc_data <- read_delim("./data/hgnc_data.txt",
delim = "\t", escape_double = FALSE, trim_ws = TRUE) %>% 
  janitor::clean_names() %>% 
  rename(target_gene = approved_symbol) %>% 
  select(-1) %>%  #remove hgnc_ids
  unique() 

#left_join
data_w_gene <- left_join(data_fin, hgnc_data, by="target_gene") %>% 
  select(cluster, target_gene, approved_name, everything())

#output the results
write.csv(data_w_gene, "./results/enrichment_analysis_results_05_12.csv")
```
